SoK: The Evolution of Defenses Against Web Application Attacks

### 2014-11-10 docs/questioning.tex
refactor discussion on ppv and npv

@@

Sensitivity and specifity are defined by way of the $2\times 2$
Table~\ref{tab:sensitivity-specificity}~\cite{linn2004}. 
In the cells of
the table we distinguish:
\begin{itemize}
\item True Positive ($a$), a real attack that raises an alarm.
\item True Negative ($d$), an event that is not an attack and that does
  not raise an alarm.
\item False Positive ($b$), an event that although it is not an attack
  raises an alarm.
\item False Negative ($c$), an event that although is as attack does
  not raise an alarm.
\end{itemize}

@@

\begin{equation}
\textrm{SE} = \textrm{Sensitivity} = \frac{a}{a + c}
\end{equation}

\begin{equation}
\textrm{SP} = \textrm{Specificity} = \frac{d}{b + d}
\end{equation}

@@

\begin{table}[ht]

\caption{Attacks and Test Outcomes in Test Environment}
\label{tab:sensitivity-specificity}

\begin{tabular}{l|l|c|c|}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No \\ \cline{2-4}
\multirow{2}{*}{Result} &  Yes &  $a = \textrm{True Positive}$ & 
$b = \textrm{False Positive}$ \\
& No & $c = \textrm{False Negative}$ & $d = \textrm{True Negative}$ \\ 
\cline{2-4}
\multicolumn{2}{r|}{Total} & $a + c$ & $b + d$ \\
\cline{3-4}
\end{tabular}

Note that the calculations are carried out vertically in
Table~\ref{tab:sensitivity-specificity}; the ratio of attack events to
normal events does not enter into sensitivity and specificity. This is
what allows us to derive them based on test data, without using data
from a real production environment. At the same time, this is what
limits their usefulness. When we obtain a test result in a production
environment, be it negative or positive, we really want to know how
much we should be worried, or relaxed. In other words, if an attack is
detected in a production environment, how much should we be worried?
If no attack is detected in a production environment, how relaxed
should we be that no attack has indeed taken place?

\begin{table}[ht]

\caption{Attacks and Test Outcomes in Real Environment}
\label{tab:ppv-npv}

\begin{tabular}{l|l|c|c|c}
\multicolumn{2}{c}{} & \multicolumn{2}{c}{Attack} & \\ \cline{3-4}
\multicolumn{2}{c|}{} & Yes & No & Total \\ \cline{2-5}
\multirow{2}{*}{Result} &  Yes &  $A = \textrm{True Positive}$ & 
$B = \textrm{False Positive}$ & \multicolumn{1}{c|}{$A + B$}\\
& No & $C = \textrm{False Negative}$ & $D = \textrm{True Negative}$ &
\multicolumn{1}{c|}{$C + D$}\\ 
\cline{2-5}
\end{tabular}

The answer to these questions is given by the \textsc{ppv} and the
\textsc{npv}. These depend not only on the test results, but also on
the \emph{prevalence} of attacks in real environments. In particular,
we can calculate them using Table~\ref{tab:ppv-npv}, which tabulates
results in a real environment, or in an environment where the
prevalence of attacks is equal to what we expect in the real world. To
indicate the difference with Table~\ref{tab:sensitivity-specificity}
we use capital letters for the entries in the table~\cite{linn2004}.
With this table, \textsc{ppv} and \textsc{npv} are given by the
following equations:

\begin{equation}
\textrm{PPV} = \frac{A}{A + C}
\end{equation}

\begin{equation}
\textrm{NPV} = \frac{D}{B + D}
\end{equation}

This time the ratios are taken horizontally; the ratio of attacks to
innocuous events affects the calculations. In fact, if PR is the
probability of attacks in the real world, the prevalence, then we
have~\cite{linn2004,altman1994}:

\begin{equation}
\textrm{PPV} = \frac{\textrm{SE}\times \textrm{PR}}{
\textrm{SE}\times \textrm{PR} + (1 - \textrm{SP})\times (1 -
\textrm{PR})}
\label{eq:ppv-se-sp}
\end{equation}

\begin{equation}
\textrm{NPV} = \frac{\textrm{SP}\times (1 - \textrm{PR})}{
(1 - \textrm{SE})\times \textrm{PR} + \textrm{SP}\times (1 -
\textrm{PR})}
\label{eq:npv-se-sp}
\end{equation}

If published research does not include figures for Sensitivity and
Specificity or values for ($a$)--($d$), on test environments, then it
is not possible to judge the accuracy of the research at all. Even if
we do have such figures, a good research tool is not necessarily a
good practical tool. Even if a research tool scores well on
Sensitivity and Specificity, it may not be very useful in the real
world if the prevalence of the caught attacks is so low that a
positive test result is not likely to indicate a real attack.

To sum up, we evaluate diagnostic performance on two criteria:

\begin{itemize}
\item Reporting of Sensitivity and Specifity, either directly, or
  indirectly from published values of $a$--$d$.
\item Reporting of \textsc{ppv} and \textsc{npv}. This can be direct
  reporting of their values, or indirect from published values of
  $A$--$D$. Also, since from equations~\ref{eq:ppv-se-sp}
  and~\ref{eq:npv-se-sp} \textsc{ppv} and \textsc{npv} can be
  calculated by the sensitity and specifity given the prevalence, it
  can be via reporting prevalance, sensitivity, specificity.
\end{itemize}

@@

\section{Lessons Learned}

In this section we point out our key findings and
discuss the lessons derived from our research.
We also argue that some of our observations may be
good reasons why the mechanisms are not adopted in practice.

One of our key findings indicate that currently,
the testing of the diagnostic performance of
mechanisms that detect top web application attacks
has not a standard form. In addition,
there are many mechanisms that are tested in a poor
manner. In many cases researchers tend not to
provide the false positive and negative results
that their tools may produce. Discussing
about the existence of such results and not
quantifying them may be a good reason to not use
such a mechanism.

A reasonable argument would be that many
mechanisms (especially the ones coming from the
etiological category), do not need to be established
purely through testing since they provide systematic
arguments as to why their design is secure against attacks.
However, implementations are not always efficient
and as we observed they can be circumvented.

>>>>

\subsection{Diagnostic Performance}
\label{ssec:diagnostic-performance}

@@

We will focus on sensitivity and specificity; we will come back to
\textsc{ppv} and \textsc{npv} in Section~\ref{sec:lessons-learned}.
Sensitivity and specifity are defined using the following~\cite{linn2004}:
\begin{itemize}
\item True Positive (\textsc{tp}), an attack that raises an alarm.
\item True Negative (\textsc{tn}), an event that is not an attack and that does
  not raise an alarm.
\item False Positive (\textsc{fp}), an event that although it is not an attack
  raises an alarm.
\item False Negative (\textsc{fn}), an event that although is as attack does
  not raise an alarm.
\end{itemize}

@@

\begin{equation}
  \textsc{se} = \textrm{Sensitivity} = \frac{\textsc{tp}}{\textsc{tp}
    + \textsc{fn}}
\label{eq:sensitivity}
\end{equation}

\begin{equation}
  \textsc{sp} = \textrm{Specificity} = \frac{\textsc{tn}}{\textsc{fp}
    + \textsc{tn}}
\label{eq:specificity}
\end{equation}

@@

@@

\section{Lessons Learned}
\label{sec:lessons-learned}

One of our key findings indicate that many detect mechanisms are
tested in a poor manner. In many cases researchers tend not to provide
the false positive and negative results that their tools may produce.
Discussing about the existence of such results and not quantifying
them also blurs the picture.

A reasonable argument would be that many mechanisms (especially the
ones coming from the etiological category), do not need to be
established purely through testing since they provide systematic
arguments as to why their design is secure against attacks. In order
for this to hold, however, an implementation of a defense mechanism
may be entirely correct in terms of its specification, which may not
happen in practice. Moreover, even a mechanism that detects an attack
based on its root cause, instead of observer behavior, may still be
circumvented so that attacks will pass undetected.

When we introduced specicifity and sensitivity in
Section~\ref{ssec:diagnostic-performance} we deferred discussion of
\textsc{ppv} and \textsc{npv} to this point. These two relate to the
performance of a detection mechanism in an actual production setting,
instead of a testbed. If an attack is detected in a production
environment, how much should we be worried? The answer is provided by
\textsc{ppv}. If no attack is detected in a production environment,
how relaxed should we be that no attack has indeed taken place? The
answer is provided by \textsc{npv}. We can calculate \textsc{ppv} and
\textsc{npv} with the following equations:

\begin{equation}
\textsc{ppv} = \frac{\textsc{tp}}{\textsc{tp} + \textsc{fp}}
\end{equation}

\begin{equation}
\textsc{npv} = \frac{\textsc{tn}}{\textsc{fn} + \textsc{tn}}
\end{equation}

Equations~\ref{eq:ppv-se-sp} and~\ref{eq:npv-se-sp} use true and false
positives and negatives, like equations~\ref{eq:sensitivity}
and~\ref{eq:specificity}. However, \textsc{tp}, \textsc{tn},
\textsc{fp}, and \textsc{fn} are not the same in the two cases:
whereas sensitivity and specificity are measured on a test
environment, \textsc{ppv} and \textsc{npv} are measured on the real,
target environment. In fact, if \textsc{pr} is the probability of
attacks in the real world, the prevalence, then we
have~\cite{linn2004,altman1994}:

\begin{equation}
\textsc{ppv} = \frac{\textsc{se}\times \textsc{pr}}{
\textsc{se}\times \textsc{pr} + (1 - \textsc{sp})\times (1 -
\textsc{pr})}
\label{eq:ppv-se-sp}
\end{equation}

\begin{equation}
\textsc{npv} = \frac{\textsc{sp}\times (1 - \textsc{pr})}{
(1 - \textsc{se})\times \textsc{pr} + \textsc{sp}\times (1 -
\textsc{pr})}
\label{eq:npv-se-sp}
\end{equation}

Of course it is not easy, and it may not even be possible, to know how
prevalent an attack is. Also, attacks on a system may depend on
factors such as its visibility and popularity, so that the same piece
of software may be subject to varying attack regimes depending on
where it is actually deployed. With this in mind it may be unfair to
ask of researchers to provide \textsc{ppv} and \textsc{npv} values for
their mechanisms---in fact, we see that nobody does.

That does not mean that deriving \textsc{ppv} and \textsc{npv} is
altogether impossible. One could deploy a system, armoured with an
attack detection mechanism, on a honeypot to study what happens over
a length of time. This could give an indication of the performance of
the mechanism out in the wild. Alternatively, one could deploy a
target, unarmoured system, on a honeypot to study the prevalence of
the class of attacks to be detected. Studying the prevalence of
classes of attacks could be an interesting area of study on its own,
which could feed directly on the evaluation of attack detection
mechanisms as practical tools.
